---
title: "Problem Set 3"
author: "Yiyun (Leo) Yao - yy3959 - (Recitation) 002 "
date: "Due Dec 1, 2023"
output:
  pdf_document: default
header-includes: 
  - \usepackage{tikz}
---

This homework must be turned in on Brightspace by Dec. 1, 2023. It must be your own work, and your own work only -- you must not copy anyone's work, or allow anyone to copy yours. This extends to writing code. You may consult with others, but when you write up, you must do so alone.

Your homework submission must be written and submitted using Rmarkdown. No handwritten solutions will be accepted. \textbf{No zip files will be accepted.} \textbf{Make sure we can read each line of code in the pdf document.} You should submit the following:

1. A compiled PDF file named yourNetID\_solutions.pdf containing your solutions to the problems.

2. A .Rmd file containing the code and text used to produce your compiled pdf named yourNetID\_solutions.Rmd.


Note that math can be typeset in Rmarkdown in the same way as Latex. Please make sure your answers are clearly structured in the Rmarkdown file:

1. Label each question part

2. Do not include written answers as code comments.

3. The code used to obtain the answer for each question part should accompany the written answer. Comment your code!


\newpage 


## Question 1 (Total: 100)
Does US military assistance strengthen or further weaken fragile and conflict-affected foreign governments? Aid may bolster state capacity and suppress violence from nonstate actors such as paramilitary groups. On the other hand, aid may be diverted to those same violent groups. To answer the question, Dube and Naidu (2015)(https://www.journals.uchicago.edu/doi/10.1086/679021?mobileUi=0) leverage changes in the allocation of US military aid to Colombian military bases. They test whether Colombian municipailites in which military bases are located have more or less paramilitary violence when the level of U.S. miliary aid increases, relative to Colombian municipalities in which miliary bases are not located. 

For this problem, you will need the `bases_replication_file.dta' file. The variables you will need are:

- parattq - DV here is paramilitary attacks
- bases6 - indicator variable whether or not there is a base in the municipality
- lrmilnar col - (logged) U.S. military and narcotics aid to Colombia
- bases6xlrmilnar col - the treatment i.e., the interaction between the level of U.S. military and narcotics aid and whether or not there is a base in the municipality
- lnnewpop - is log of population

### Part a (60 points)

The treatment in this case is a continuous 'intensity' variable that changes over time. The authors use the interaction between the level of U.S. military and narcotics aid and whether a base exists in a municipality. How many units are in the 'control' group (no bases)? Does the bases variable change over time or is it a unit-constant factor? How about the logged military aid variable, does it change across units for a given year? What do the authors seem to be assuming about how military aid is allocated?

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(haven)
library(estimatr) # for lm with robust se : ?lm_robust()

# Load bases data
bases <- haven::read_dta("bases_replication_final.dta")

# How many observations are in the ``no bases group"
no_base <- bases %>%
  filter(bases6 == 0)
no_base_num <- no_base %>% nrow()
no_base_num

```
There are 16272 observations in the control group (no bases).
```{r}
## How about each of them?

bases6_count_total <- bases %>%
  group_by(year) %>%
  summarise(count_total = n(), unique_bases6_count_total = n_distinct(bases6), 
            .groups = 'drop')
bases6_count_total

```
```{r}

bases6_count_control <- no_base %>%
  group_by(year) %>%
  summarise(count_control = n(), unique_bases6_count_control = n_distinct(bases6), 
            .groups = 'drop')
bases6_count_control

```
The bases variable is constant in time. It does not change over time.
```{r}
logaid_count_total <- bases %>%
  group_by(municipality) %>%
  summarise(count_total = n(), .groups = 'drop')
logaid_count_total

```
```{r}
logaid_count_control <- no_base %>%
  group_by(municipality) %>%
  summarise(count_control = n(), .groups = 'drop')
logaid_count_control

```
The logged military aid variable is constant in unit. It does not change across units for a given year.
```{r}

## How many municipalities do we have
municipalities_count_total <- bases %>%
  group_by(year) %>%
  summarise(count_total = n_distinct(municipality), .groups = 'drop')
municipalities_count_total

```
```{r}

## How many municipalities do we have
municipalities_count_control <- no_base %>%
  group_by(year) %>%
  summarise(count_control = n_distinct(municipality), .groups = 'drop')
municipalities_count_control

```
In the control group (no base), we have 904 municipalities. Overall, we have 936 municipalities.

The authors seem to be assuming that military aid is allocated based on the presence of military bases in the municipalities. From the results above, I'm inferring that the author is assuming the military aid is consistent across the years for each municipality and that the aid is uniformly distributed for each municipality within the same year.

### Part b (20 points)
The authors use a common empirical strategy called two-way fixed effects to estimate the average treatment effect of military aid. The model they estimate includes fixed effects for both time periods and units (and includes logged population as an additional covariate):

$$ Y_{it} = \gamma_t + \alpha_i + \tau D_{it} + \beta X_{it} + \epsilon_{it}$$ 
What assumptions are the authors making in order to identify the treatment effect of military aid?

1. SUTVA
2. Ignorability
3. Unconfoundedness
4. Time-invariant unobserved heterogeneity / Unobserved confounding is constant within each group.
For example, any unobserved factors that could potentially influence both the treatment assignment and the outcome are assumed to be constant over time. Another assumption that also lies in this category would be that, for each municipality/year, there are unobserved factors that remain constant over time and are unique to each municipality/year, and these factors would affect the outcome. This helps control for unobserved factors that are constant within each municipality and do not change over time. 
5. Parallel trends / Common trends
6. Outcome model is linear / Independence of errors

### Part c (20 points)

Using the two-way fixed effects estimator, estimate the effect of U.S. military and narcotics aid on the number of paramilitary attacks, including log of population as a covariate. The two sets of fixed effects are for municipality (municipality) and year (year). Cluster your standard errors at the unit level (see the cluster argument in lm_robust. Report a 95% confidence interval for your estimate and interpret your results.

```{r}
#?lm_robust (set se_type to "CR0")
# Fit Regression using lm_robust 
model <- tidy(lm_robust(paratt ~ bases6xlrmilnar_col + lnnewpop 
                        + factor(municipality) + factor(year),
                        data = bases, clusters = municipality, se_type = 'CR0'))

# Summarize
summary(model)

```
```{r}
# Extracting the treatment effect
model[2, ]
```
From the summary of the results, we can see that municipalities, time, and population all have different impacts on the paramilitary attacks. The estimate for the treatment effect is 0.15 at a 95% confidence interval [0.032, 0.268]. The standard error is 0.06, which is very small. This estimate of 0.15 means that the U.S. military and narcotics aid increases the average number of paramilitary attacks by 0.15, including log of population as a covariate. This shows the positive relationship between military aids and number paramilitary attacks. More aids causes higher level of paramilitary violence.
